Best Practices for Developing Monte Carlo Methodologies in Molecular Simulations [Article v\versionnumber]

### 2025-06-02 paper/paper.tex
updated version and some space tildes

\newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
% Do not add a newline in the next command, no matter how long the repository name is, as it will break the link in the PDF.
\newcommand{\githubrepository}{\url{https://github.com/usnistgov/best-practices-mc}}  %this should be the main github repository for this article.

@@

The derivation of $\chi$ in this article requires the condition of detailed balance, shown in Fig.~\ref{fig:detailed_balance}, for a transition from an old microstate, $o$ to a new microstate, $n$,
\begin{equation}
\Pi_o\alpha_{o\rightarrow n}\pi_{o\rightarrow n} = \Pi_n\alpha_{n\rightarrow o}\pi_{n\rightarrow o},
\label{eq:detailed_balance_first}
\end{equation}
where $\Pi_o$ and $\Pi_n$ are the probabilities of being in the old and new microstates, respectively, $\alpha_{o\rightarrow n}$ and $\alpha_{n\rightarrow o}$ are the probabilities of attempting a transition from the old to the new microstate, and vice versa, and $\pi_{o\rightarrow n}$ and $\pi_{n\rightarrow o}$ are the probabilities of accepting those transitions.
Symbols in this article are chosen to follow closely with Ref. \cite{frenkel_understanding_2002}, but note the different definition of $\pi$.
Detailed balance is a sufficient, but not necessary, condition for ergodic sampling \cite{manousiouthakis_strict_1999}.
For simplicity, we use detailed balance throughout the derivations of this article.
While the microstate probabilities, $\Pi_o$ and $\Pi_n$, are determined solely by the statistical mechanical ensemble, the transition probabilities $\alpha_{o\rightarrow n}$ and $\alpha_{n\rightarrow o}$ depend upon how the MC trial move is implemented.
Thus, Eq.~\ref{eq:detailed_balance_first} provides the route to obtain the trial acceptance probability after the ensemble, trial implementation and acceptance criterion are defined.
Here, \emph{trial implementation} encompasses a great variety of potential moves, such as displacements, volume changes, temperature changes, particle insertions and deletions, alchemical transformations, and more.

@@

Eq.~\ref{eq:lhs_dv} is not valid for uniformly random changes in $\ln V$.
Specifically, uniform changes in $\ln V$ occur when the particle coordinates and periodic boundary lengths are scaled by a factor of $e^{(\delta\ln V)/D}$, where $\delta\ln V$ is chosen as a random amount selected uniformly within $\pm\delta\ln V_{\mathrm{max}}$, and $\Delta V = V_o(e^{\delta\ln V}-1)$.
As discussed in Ref. \cite{frenkel_understanding_2002} (without a shell particle) and Ref. \cite{corti_monte_2002} and Section~\ref{sec:rhs_npt} (with a shell particle), uniform changes in $\ln V$ use Eq.~\ref{eq:rhs_npt_delta_lnv} instead of Eq.~\ref{eq:rhs_npt_delta_v}, resulting in the following acceptance probability
\begin{equation}
\chi=\left(\frac{V_n}{V_o}\right)^{N+1}e^{-\beta(\Delta U + P\Delta V)}.
\label{eq:lhs_dlnv}
\end{equation}
Code Block~\ref{sim:ideal_gas_npt} uses Eq.~\ref{eq:lhs_dv} or Eq.~\ref{eq:lhs_dlnv} for an ideal gas to compare against theory for uniform changes in $V$ or $\ln V$, respectively.

@@

If a shell particle is introduced to remove redundant microstates in the partition function~\cite{attard_density_1995, koper_length_1996, corti_deriving_1998, han_isothermal-isobaric_2001, corti_isothermal-isobaric_2001, corti_monte_2002, stroker_systematic_2021, hatch_theory_2024}, then there is one less particle that is scaled by the volume and Eqs.~\ref{eq:rhs_npt_pi}, \ref{eq:rhs_npt_delta_v} and \ref{eq:rhs_npt_delta_lnv} become 
\begin{equation}
\Pi \diff\mathbf{s}^{N-1}\diff\boldsymbol{\omega}^{N}\diff V \propto V^{N-1} e^{-\beta(U+P V)} \diff\mathbf{s}^{N-1} \diff\boldsymbol{\omega}^{N}\diff V,
\end{equation}
\begin{equation}
\frac{\Pi_n}{\Pi_o} = \left(\frac{V_n}{V_o}\right)^{N-1}e^{-\beta(\Delta U + P\Delta V)},
\end{equation}
and
\begin{equation}
\frac{\Pi_n}{\Pi_o} = \left(\frac{V_n}{V_o}\right)^{N}e^{-\beta(\Delta U + P\Delta V)},
\end{equation}
respectively.
The acceptance probabilities shown in Section \ref{sec:lhs_dv} without a shell particle would change according to these new microstate probabilities when a shell particle is included.
Specifically, Eqs. \ref{eq:lhs_dv} and \ref{eq:lhs_dlnv} become
\begin{equation}
\chi=\left(\frac{V_n}{V_o}\right)^{N-1}e^{-\beta(\Delta U + P\Delta V)}.
\end{equation}
and
\begin{equation}
\chi=\left(\frac{V_n}{V_o}\right)^{N}e^{-\beta(\Delta U + P\Delta V)},
\end{equation}
for uniform changes in $V$ and $\ln V$, respectively.

@@

Consider an alternative approach that includes the indistinguishable particle $N!$ correction, and uses scaled coordinates, as described in Ref. \cite{frenkel_understanding_2002}.
In this case, the $\mu VT$ microstate probability is
\begin{equation}
\Pi(N_i)\diff\mathbf{s}^{N}\diff\boldsymbol{\omega}^{N} \propto \frac{V^{N}z_i^{N_i}}{N!} e^{-\beta U} \diff\mathbf{s}^{N}\diff\boldsymbol{\omega}^{N}.
\end{equation}
The ratio of $\mu VT$ microstate probabilities for the special case of indistinguishable particles and scaled coordinates is then
\begin{equation}
\frac{\Pi_n}{\Pi_o} = \frac{N_{io}!}{N_{in}!}\left(V z_i\diff\mathbf{s}\diff\boldsymbol{\omega}\right)^{\Delta N_i}e^{-\beta\Delta U},
\label{eq:rhs_muvt_alt}
\end{equation}
and is valid for changing the number of indistinguishable particles and choosing a newly inserted particle position in scaled coordinates, $\mathbf{s}$, not $\mathbf{r}$.

@@

As opposed to deriving insertion and deletion acceptances with distinguishable particles in non-scaled coordinates, as in Sections~\ref{sec:rhs_muvt_distinguishable} and \ref{sec:lhs_insdel}, consider a similar trial conducted with indistinguishable particles in scaled coordinates, as in Ref. \cite{frenkel_understanding_2002}.
In this case, the first ratio in Eq.~\ref{eq:detailed_balance} is given by Eq.~\ref{eq:rhs_muvt_alt}.
But since the transition probabilities also change, the same acceptances are nonetheless obtained.

@@

Using Table~\ref{tab:lhs_ins_cavity_testpoint} and substituting Eq.~\ref{eq:rhs_muvt} into Eq.~\ref{eq:detailed_balance}, the acceptance probability for cavity bias particle insertion with $N_{test}$ points is
\begin{equation}
\chi = N_l \delta^D\left(\frac{n_c}{N_{test}}\right)\frac{z_i}{P_{\omega n}(N_i+1)}e^{-\beta\Delta U}.
\label{eq:cavity_bias_ntest}
\end{equation}
This equation takes the form of Eq. 4 of Ref. \cite{snurr_prediction_1993} because the fraction of the points in the cavity is $P_c = n_c/N_{test}$ and the total volume considered for insertions is $v=N_l\delta^D$.

@@

Using Table~\ref{tab:lhs_disp_in_avb2} and substituting Eq.~\ref{eq:rhs_nvt} into Eq.~\ref{eq:detailed_balance}, the acceptance probability is
\begin{equation}
\chi = \frac{P_{\omega o}(1-P_{bias})(N_i-N_i^{AV}-\delta_{ij})v_{AV}}{P_{\omega n} P_{bias}(N_i^{AV}+1)(V-v_{AV})}e^{-\beta \Delta U}.
\label{eq:avb2outin}
\end{equation}
For an example of combining AVB with CB, see the Appendix of Ref. \cite{hatch_self-assembly_2016}.

@@

Although the derivation of the acceptance criteria for MC trials is the focus of this article, common errors encountered during the implementation of MC are included here.
For a more comprehensive discussion of the inner workings of MC see Ref. \cite{dubbeldam_inner_2013}.
Thus, the scope of this section is for testing MC trial implementations.
In this section, we discuss energy tests and machine precision issues for helping to capture common errors, the calculation of standard deviations using the block method \cite{flyvbjerg_error_1989, grossfield_best_2018}, the generation of a random position on the surface of a unit sphere and spherical shell, and the generation of random quaternions and rotation matrices.
This section then ends with a discussion of the general workflow for testing a new MC trial.

>>>>

\newcommand{\versionnumber}{1.0}  % you should update the minor version number in preprints and major version number of submissions.
% Do not add a newline in the next command, no matter how long the repository name is, as it will break the link in the PDF.
\newcommand{\githubrepository}{\url{https://github.com/usnistgov/best-practices-mc}}  %this should be the main github repository for this article.

@@

The derivation of $\chi$ in this article requires the condition of detailed balance, shown in Fig.~\ref{fig:detailed_balance}, for a transition from an old microstate, $o$ to a new microstate, $n$,
\begin{equation}
\Pi_o\alpha_{o\rightarrow n}\pi_{o\rightarrow n} = \Pi_n\alpha_{n\rightarrow o}\pi_{n\rightarrow o},
\label{eq:detailed_balance_first}
\end{equation}
where $\Pi_o$ and $\Pi_n$ are the probabilities of being in the old and new microstates, respectively, $\alpha_{o\rightarrow n}$ and $\alpha_{n\rightarrow o}$ are the probabilities of attempting a transition from the old to the new microstate, and vice versa, and $\pi_{o\rightarrow n}$ and $\pi_{n\rightarrow o}$ are the probabilities of accepting those transitions.
Symbols in this article are chosen to follow closely with Ref.~\cite{frenkel_understanding_2002}, but note the different definition of $\pi$.
Detailed balance is a sufficient, but not necessary, condition for ergodic sampling \cite{manousiouthakis_strict_1999}.
For simplicity, we use detailed balance throughout the derivations of this article.
While the microstate probabilities, $\Pi_o$ and $\Pi_n$, are determined solely by the statistical mechanical ensemble, the transition probabilities $\alpha_{o\rightarrow n}$ and $\alpha_{n\rightarrow o}$ depend upon how the MC trial move is implemented.
Thus, Eq.~\ref{eq:detailed_balance_first} provides the route to obtain the trial acceptance probability after the ensemble, trial implementation and acceptance criterion are defined.
Here, \emph{trial implementation} encompasses a great variety of potential moves, such as displacements, volume changes, temperature changes, particle insertions and deletions, alchemical transformations, and more.

@@

Eq.~\ref{eq:lhs_dv} is not valid for uniformly random changes in $\ln V$.
Specifically, uniform changes in $\ln V$ occur when the particle coordinates and periodic boundary lengths are scaled by a factor of $e^{(\delta\ln V)/D}$, where $\delta\ln V$ is chosen as a random amount selected uniformly within $\pm\delta\ln V_{\mathrm{max}}$, and $\Delta V = V_o(e^{\delta\ln V}-1)$.
As discussed in Ref.~\cite{frenkel_understanding_2002} (without a shell particle) and Ref.~\cite{corti_monte_2002} and Section~\ref{sec:rhs_npt} (with a shell particle), uniform changes in $\ln V$ use Eq.~\ref{eq:rhs_npt_delta_lnv} instead of Eq.~\ref{eq:rhs_npt_delta_v}, resulting in the following acceptance probability
\begin{equation}
\chi=\left(\frac{V_n}{V_o}\right)^{N+1}e^{-\beta(\Delta U + P\Delta V)}.
\label{eq:lhs_dlnv}
\end{equation}
Code Block~\ref{sim:ideal_gas_npt} uses Eq.~\ref{eq:lhs_dv} or Eq.~\ref{eq:lhs_dlnv} for an ideal gas to compare against theory for uniform changes in $V$ or $\ln V$, respectively.

@@

If a shell particle is introduced to remove redundant microstates in the partition function~\cite{attard_density_1995, koper_length_1996, corti_deriving_1998, han_isothermal-isobaric_2001, corti_isothermal-isobaric_2001, corti_monte_2002, stroker_systematic_2021, hatch_theory_2024}, then there is one less particle that is scaled by the volume and Eqs.~\ref{eq:rhs_npt_pi}, \ref{eq:rhs_npt_delta_v} and \ref{eq:rhs_npt_delta_lnv} become 
\begin{equation}
\Pi \diff\mathbf{s}^{N-1}\diff\boldsymbol{\omega}^{N}\diff V \propto V^{N-1} e^{-\beta(U+P V)} \diff\mathbf{s}^{N-1} \diff\boldsymbol{\omega}^{N}\diff V,
\end{equation}
\begin{equation}
\frac{\Pi_n}{\Pi_o} = \left(\frac{V_n}{V_o}\right)^{N-1}e^{-\beta(\Delta U + P\Delta V)},
\end{equation}
and
\begin{equation}
\frac{\Pi_n}{\Pi_o} = \left(\frac{V_n}{V_o}\right)^{N}e^{-\beta(\Delta U + P\Delta V)},
\end{equation}
respectively.
The acceptance probabilities shown in Section \ref{sec:lhs_dv} without a shell particle would change according to these new microstate probabilities when a shell particle is included.
Specifically, Eqs.~\ref{eq:lhs_dv} and \ref{eq:lhs_dlnv} become
\begin{equation}
\chi=\left(\frac{V_n}{V_o}\right)^{N-1}e^{-\beta(\Delta U + P\Delta V)}.
\end{equation}
and
\begin{equation}
\chi=\left(\frac{V_n}{V_o}\right)^{N}e^{-\beta(\Delta U + P\Delta V)},
\end{equation}
for uniform changes in $V$ and $\ln V$, respectively.

@@

Consider an alternative approach that includes the indistinguishable particle $N!$ correction, and uses scaled coordinates, as described in Ref.~\cite{frenkel_understanding_2002}.
In this case, the $\mu VT$ microstate probability is
\begin{equation}
\Pi(N_i)\diff\mathbf{s}^{N}\diff\boldsymbol{\omega}^{N} \propto \frac{V^{N}z_i^{N_i}}{N!} e^{-\beta U} \diff\mathbf{s}^{N}\diff\boldsymbol{\omega}^{N}.
\end{equation}
The ratio of $\mu VT$ microstate probabilities for the special case of indistinguishable particles and scaled coordinates is then
\begin{equation}
\frac{\Pi_n}{\Pi_o} = \frac{N_{io}!}{N_{in}!}\left(V z_i\diff\mathbf{s}\diff\boldsymbol{\omega}\right)^{\Delta N_i}e^{-\beta\Delta U},
\label{eq:rhs_muvt_alt}
\end{equation}
and is valid for changing the number of indistinguishable particles and choosing a newly inserted particle position in scaled coordinates, $\mathbf{s}$, not $\mathbf{r}$.

@@

As opposed to deriving insertion and deletion acceptances with distinguishable particles in non-scaled coordinates, as in Sections~\ref{sec:rhs_muvt_distinguishable} and \ref{sec:lhs_insdel}, consider a similar trial conducted with indistinguishable particles in scaled coordinates, as in Ref.~\cite{frenkel_understanding_2002}.
In this case, the first ratio in Eq.~\ref{eq:detailed_balance} is given by Eq.~\ref{eq:rhs_muvt_alt}.
But since the transition probabilities also change, the same acceptances are nonetheless obtained.

@@

Using Table~\ref{tab:lhs_ins_cavity_testpoint} and substituting Eq.~\ref{eq:rhs_muvt} into Eq.~\ref{eq:detailed_balance}, the acceptance probability for cavity bias particle insertion with $N_{test}$ points is
\begin{equation}
\chi = N_l \delta^D\left(\frac{n_c}{N_{test}}\right)\frac{z_i}{P_{\omega n}(N_i+1)}e^{-\beta\Delta U}.
\label{eq:cavity_bias_ntest}
\end{equation}
This equation takes the form of Eq.~4 of Ref.~\cite{snurr_prediction_1993} because the fraction of the points in the cavity is $P_c = n_c/N_{test}$ and the total volume considered for insertions is $v=N_l\delta^D$.

@@

Using Table~\ref{tab:lhs_disp_in_avb2} and substituting Eq.~\ref{eq:rhs_nvt} into Eq.~\ref{eq:detailed_balance}, the acceptance probability is
\begin{equation}
\chi = \frac{P_{\omega o}(1-P_{bias})(N_i-N_i^{AV}-\delta_{ij})v_{AV}}{P_{\omega n} P_{bias}(N_i^{AV}+1)(V-v_{AV})}e^{-\beta \Delta U}.
\label{eq:avb2outin}
\end{equation}
For an example of combining AVB with CB, see the Appendix of Ref.~\cite{hatch_self-assembly_2016}.

@@

Although the derivation of the acceptance criteria for MC trials is the focus of this article, common errors encountered during the implementation of MC are included here.
For a more comprehensive discussion of the inner workings of MC see Ref.~\cite{dubbeldam_inner_2013}.
Thus, the scope of this section is for testing MC trial implementations.
In this section, we discuss energy tests and machine precision issues for helping to capture common errors, the calculation of standard deviations using the block method \cite{flyvbjerg_error_1989, grossfield_best_2018}, the generation of a random position on the surface of a unit sphere and spherical shell, and the generation of random quaternions and rotation matrices.
This section then ends with a discussion of the general workflow for testing a new MC trial.
